\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{enumitem}

\definecolor{codebg}{RGB}{245,245,250}
\definecolor{cmdbg}{RGB}{40,42,54}
\definecolor{cmdfg}{RGB}{248,248,242}

\lstdefinestyle{py}{
    language=Python,
    basicstyle=\ttfamily\small,
    backgroundcolor=\color{codebg},
    keywordstyle=\bfseries\color{blue!70!black},
    stringstyle=\color{green!50!black},
    commentstyle=\itshape\color{gray},
    breaklines=true, frame=single, framesep=4pt, showstringspaces=false}

\lstdefinestyle{shell}{
    language=bash,
    basicstyle=\ttfamily\footnotesize\color{cmdfg},
    backgroundcolor=\color{cmdbg},
    breaklines=true, frame=single, framesep=4pt, showstringspaces=false}

\hypersetup{colorlinks=true, urlcolor=blue, linkcolor=black,
            citecolor=black, breaklinks=true}

\title{\Large\bfseries triloop \& picoacq\\[2pt]
       \large Analysis Code Reference\\[8pt]
       \normalsize Pipeline math, module-by-module walkthrough, and
       per-program description}
\author{}
\date{\today \\[6pt] \small Version 0.1.0}

\begin{document}
\maketitle

\tableofcontents
\bigskip\hrule\bigskip

%-----------------------------------------------------------------------
\section{Introduction}
%-----------------------------------------------------------------------

This document is a developer's reference for the analysis code in
\textsf{triloop} and the acquisition code in \textsf{picoacq}.  It
complements the \textit{triloop User Manual} (which describes how to
\emph{use} the tools) and the \textit{picoacq User Manual} (which
describes the hardware and the capture file format).  Here we focus on
\emph{what each Python module does}, in what order they are called by
the public entry points, and the mathematics each step implements.

\paragraph{Audience.} Reader is comfortable with Python, complex
baseband signal processing, and standard linear-algebra notation.

\paragraph{Conventions used in this document.} Bold lower-case
($\mathbf b$) denotes a vector; bold upper-case ($\mathsf{N}$) denotes
a matrix; $\hat{\mathbf k}$ is the unit propagation vector \emph{toward
the source}; angles are in degrees in code, radians in equations.

%-----------------------------------------------------------------------
\section{Pipeline overview}
%-----------------------------------------------------------------------

A capture flows through the toolchain in four stages:

\begin{enumerate}\itemsep=2pt
\item \textbf{Acquisition} (\textsf{picoacq}).  Hardware capture
      orchestrated by \texttt{picoacq.recorder.capture()}, which calls
      the PicoSDK wrapper \texttt{picoacq.ps5444a} or, in
      \texttt{--simulate} mode, \texttt{picoacq.simulator}.  Output:
      a triloop-format HDF5 file with float32 samples in volts plus a
      JSON \texttt{capture\_settings} dict carrying the chosen sample
      rate, voltage ranges, auto-range probe values, and (new in v0.1)
      the list of expected RF bands \texttt{rf\_bands\_hz}.
\item \textbf{Band extraction} (\texttt{triloop.bands},
      \texttt{triloop.extract}).  Read the HDF5 file with
      \texttt{triloop.io\_hdf5.read\_capture()}; identify the baseband
      alias position of each RF band via \texttt{picoacq.alias};
      mix-and-LPF each channel down to a complex baseband centred on
      that band; decimate to a workable IF rate.
\item \textbf{Polarimetry and direction finding}
      (\texttt{triloop.analyze}, \texttt{triloop.geometry},
      \texttt{triloop.stokes}, \texttt{triloop.direction},
      \texttt{triloop.beamform}).  Recover the lab-frame
      $\mathbf B$-vector from the three loops via the inverse of the
      geometry matrix $\mathsf{N}$, project onto the plane
      perpendicular to $\hat{\mathbf k}$, compute Stokes parameters
      and intensity, and (optionally) refine $\hat{\mathbf k}$ with
      a null-eigenvector / null-sweep search.
\item \textbf{Reporting} (\texttt{triloop.view},
      \texttt{triloop.multiband}, \texttt{triloop.browse}).  Static
      QC plots, multi-band comparison figures, and an interactive
      single-file HTML browser for visual inspection.
\end{enumerate}

The CLI tools (\texttt{triloop view}, \texttt{triloop analyze-multi},
\texttt{triloop browse}, \texttt{triloop analyze},
\texttt{triloop locate}, \texttt{triloop batch}, \texttt{triloop
summary}, \texttt{triloop simulate}, \texttt{triloop notebook}) wrap
these stages in different combinations.

%-----------------------------------------------------------------------
\section{Acquisition: \textsf{picoacq}}
%-----------------------------------------------------------------------

\subsection{\texttt{picoacq.cli}}

Click-based command-line interface with two subcommands:
\texttt{capture} (one-shot) and \texttt{monitor} (continuous).  Each
exposes the same configuration knobs as the underlying recorder; the
defaults reflect the multi-band undersampling workflow:

\begin{itemize}\itemsep=2pt
\item \texttt{--rate 15625000} (Hz)
\item \texttt{--duration 8.0} (s)
\item \texttt{--channels A,B,C,D}
\item \texttt{--range 2.0} (V; only used when \texttt{--no-auto-range})
\item \texttt{--auto-range/--no-auto-range} (default on)
\item \texttt{--headroom 3.0}
\item \texttt{--probe 0.1} (auto-range probe duration, s)
\item \texttt{--rf-bands "2.5e6,5e6,10e6,15e6,20e6"}
\end{itemize}

\subsection{\texttt{picoacq.ps5444a}}

A thin \texttt{ctypes} wrapper around the PicoSDK 5000a driver.  Five
functions:

\begin{itemize}\itemsep=2pt
\item \texttt{open\_5444(n\_channels=4)} -- context manager that opens
      the device, automatically picking 16-bit resolution for $\le 2$
      channels and 14-bit for 3--4 channels (5444D constraint).
\item \texttt{configure\_channel(ps, handle, ch, range\_volts,
      coupling)} -- enables a channel and sets its $\pm$range.
\item \texttt{get\_timebase(ps, handle, target\_rate)} -- searches the
      Pico timebase index space for the closest valid sample rate at
      or below the requested target.  Skips early indices that are
      invalid in 4-channel mode.
\item \texttt{acquire\_block(ps, handle, channels, n, tb, ranges)} --
      block-mode acquisition with explicit ctypes-wrapped arguments
      to avoid signature mismatches between picosdk versions.
\item \texttt{\_ensure\_pico\_dyld\_path()} -- macOS-only shim that
      injects the framework library path into
      \texttt{DYLD\_LIBRARY\_PATH} before importing \texttt{picosdk}.
\end{itemize}

\subsection{\texttt{picoacq.simulator}}

The no-hardware fallback.  Generates a 3-loop synthetic dataset with
configurable arrival direction, polarization, additive Gaussian noise,
and Faraday rotation rate.  Used by \texttt{--simulate} mode and as
the silent fallback when the SDK or hardware is unavailable.

\subsection{\texttt{picoacq.recorder}}

The orchestrator.  Top-level function:
\texttt{capture(output\_file, duration\_s, sample\_rate, channels,
ranges\_volts, coupling, loops\_config, simulate, sim\_kwargs,
auto\_range\_enabled, headroom, probe\_duration\_s, rf\_bands,
verbose)}.  Internally:

\begin{enumerate}\itemsep=2pt
\item \texttt{\_check\_buffer()} validates that
      $\mathrm{duration}\times\mathrm{rate}$ samples fit in the
      onboard memory (128\,MS/ch in 4-channel mode); raises
      \texttt{ValueError} with a clear maximum if not.
\item If auto-range is on (and \texttt{ranges\_volts} not supplied):
      open the scope at the maximum $\pm 20$\,V on every channel,
      run a brief probe capture (\texttt{auto\_range()}), measure
      peak/RMS in volts, and call \texttt{\_pick\_range()} to choose
      the smallest available Pico range whose full scale
      $\geq K\cdot$peak (default $K = 3$).
\item Run the science capture via \texttt{\_try\_real\_capture()};
      on any SDK/hardware failure, fall back transparently to
      \texttt{simulate\_capture()}.
\item Convert int16 ADC codes $\to$ float32 volts; write to HDF5 via
      \texttt{triloop.io\_hdf5.write\_capture()} with the chosen
      ranges, the auto-range probe results, and \texttt{rf\_bands\_hz}
      stored in \texttt{capture\_settings}.
\end{enumerate}

\subsection{\texttt{picoacq.alias}}
\label{sec:alias}

Pure-Python helper that maps RF frequencies to baseband alias positions
for a given sample rate.  An RF tone at $f_{\rm rf}$ folds to baseband as
\begin{equation}
f_{\rm bb} = \bigl|\,f_{\rm rf} - n\,f_s\,\bigr|,
\qquad n = \mathrm{round}(f_{\rm rf}/f_s),
\end{equation}
with Nyquist zone $z = \lfloor 2 f_{\rm rf}/f_s\rfloor + 1$.  Even
zones produce a frequency-inverted spectrum.  Three exports:

\begin{itemize}\itemsep=2pt
\item \texttt{Alias} dataclass with fields
      \texttt{(rf\_hz, baseband\_hz, nyquist\_zone, inverted)}.
\item \texttt{alias\_of(rf\_hz, sample\_rate\_hz)} -- single mapping.
\item \texttt{rf\_for\_baseband(f\_bb, rf\_bands, fs, tol)} --
      inverse lookup (given a detected baseband peak, return which
      RF band it most likely came from).
\end{itemize}

This module is the single source of truth for alias bookkeeping;
\texttt{triloop.bands} and \texttt{triloop.view}/\texttt{browse} all
import from it.

%-----------------------------------------------------------------------
\section{The capture file format}
\label{sec:fileformat-ref}
%-----------------------------------------------------------------------

Both packages read and write the same HDF5 v0.1 layout:

\begin{lstlisting}[style=shell]
/                           HDF5 root, @format_version = "0.1"
/metadata                   group, attributes:
    @sample_rate            float64, Hz
    @duration_s             float64
    @start_time_utc         ISO-8601
    @scope_model            string
    @scope_serial           string
    @capture_settings       JSON-encoded dict
    @loops_config           JSON-encoded dict
/channels
    /A   /B   /C   /D       float32 (or int16) datasets, length N
    each has:
        @gain_v_per_count   float64
        @offset_v           float64
/time                       optional float64 array, length N
\end{lstlisting}

\noindent\texttt{capture\_settings} fields recorded by
\texttt{picoacq.recorder.capture()}:
\begin{itemize}\itemsep=2pt
\item \texttt{requested\_duration\_s, requested\_sample\_rate}
\item \texttt{channels, ranges\_volts, coupling}
\item \texttt{used\_simulator, auto\_range\_enabled, headroom}
\item \texttt{auto\_range\_probe} (\texttt{peaks\_volts, rms\_volts,
      headroom, probe\_duration\_s}) -- only when probe ran
\item \texttt{picoscope\_ranges\_volts} -- the actual chosen ranges
\item \texttt{rf\_bands\_hz} -- list of expected RF frequencies in Hz
\item \texttt{truth} -- only present in simulator mode
\end{itemize}

\texttt{triloop.io\_hdf5} provides \texttt{read\_capture()} and
\texttt{write\_capture()}; both are short and tested, and any third
party can interoperate by writing the same layout.

%-----------------------------------------------------------------------
\section{Geometry and direction transforms}
%-----------------------------------------------------------------------

\subsection{\texttt{triloop.geometry}}

Operates on the loop-array geometry config dict
(\texttt{loops\_config}).  Five public functions:

\begin{itemize}\itemsep=2pt
\item \texttt{az\_el\_to\_khat(az\_deg, el\_deg)} returns
      $\hat{\mathbf k} = (\cos\theta\sin\varphi, \cos\theta\cos\varphi,
      \sin\theta)^\top$ in the local ENU frame.
\item \texttt{build\_N\_matrix(loops\_config)} returns the $3\times 3$
      matrix $\mathsf{N}$ whose $i$-th row is the $i$-th loop's normal
      $\hat{\mathbf n}_i$ (in ENU, scaled by per-loop gain).  This is
      the matrix that maps the lab-frame $\mathbf B$ to the
      loop-projected real signal $\mathbf b = \mathsf{N}\mathbf B$.
\item \texttt{perp\_projector(khat)} returns
      $\mathsf{P}_\perp = \mathsf{I} - \hat{\mathbf k}\hat{\mathbf k}^\top$,
      the projector onto the plane perpendicular to $\hat{\mathbf k}$.
\item \texttt{perp\_orthonormal\_basis(khat)} returns two unit vectors
      $(\hat{\mathbf p}, \hat{\mathbf q})$ that span that plane, with
      $\hat{\mathbf p}$ chosen along the local horizontal whenever
      possible, $\hat{\mathbf q} = \hat{\mathbf k}\times\hat{\mathbf p}$.
\item \texttt{LOOP\_NORMALS\_DEFAULT} and \texttt{\_normal\_from\_az\_el}
      support the default cube-vertex layout.
\end{itemize}

The default geometry is three loops on three faces of a cube whose
body diagonal is vertical, giving normal-tilts of $\arctan(1/\sqrt{2})
\approx 35.26^\circ$ above horizontal at azimuths $\{0, +120, -120\}^\circ$.

%-----------------------------------------------------------------------
\section{Band extraction and analysis}
%-----------------------------------------------------------------------

\subsection{\texttt{triloop.extract}}

Two functions covering the single-band path:

\begin{itemize}\itemsep=2pt
\item \texttt{extract\_complex\_baseband(t, b, f0, BW)} performs an
      FFT-based peak search around $f_0$ over a $\pm$\texttt{search\_BW}
      window, mixes the strongest peak to DC, brick-wall low-passes to
      $\pm$\texttt{BW}/2, and returns the complex baseband at the
      original sample rate, the detected peak frequency, and an
      FFT-domain SNR estimate.
\item \texttt{extract\_three\_loops(...)} calls the above on each loop,
      applies per-loop \texttt{phase\_offset\_deg} calibration, and
      returns a stacked $(3, N)$ complex array.
\end{itemize}

\subsection{\texttt{triloop.bands}}

Multi-band counterpart, used when the capture's
\texttt{rf\_bands\_hz} populates more than one entry.  Public
function: \texttt{extract\_bands(cap, rf\_bands, bw\_hz, decim\_rate\_hz, channels)}.
For every band:

\begin{enumerate}\itemsep=2pt
\item \texttt{picoacq.alias.alias\_of()} computes
      $f_{\rm bb}$, the Nyquist zone, and \texttt{inverted}.
\item Validate that the requested \texttt{bw\_hz} doesn't reach into
      the adjacent Nyquist edge or another band's alias.
\item Mix every selected channel down to $f_{\rm bb}$ via FFT
      (\texttt{\_extract\_one\_channel}), brick-wall filter to
      $\pm$\texttt{bw\_hz}/2, decimate to the smallest integer factor
      that yields $\geq$ \texttt{decim\_rate\_hz}, and conjugate the
      result if the band lives in an even (inverted) Nyquist zone.
\end{enumerate}

Returns a dict mapping \texttt{rf\_hz} $\to$ \texttt{BandExtraction}
dataclass with fields
\texttt{rf\_hz, baseband\_hz, nyquist\_zone, inverted, bw\_hz,
decim\_rate\_hz, t, z}, where \texttt{z} is a per-channel dict of
\texttt{complex64} arrays.

\subsection{\texttt{triloop.analyze}}

The polarimetry pipeline.  Two public entry points:

\begin{itemize}\itemsep=2pt
\item \texttt{analyze(t, B1, B2, B3, f0, BW, az\_deg, el\_deg, loops\_config)}
      -- starts from real loop signals, calls
      \texttt{extract\_three\_loops()}, then forwards to:
\item \texttt{analyze\_z\_loops(t, Z\_loops, f\_peak, snrs\_db, az\_deg, el\_deg, loops\_config)}
      -- the inner pipeline, also called directly by the multi-band
      path.
\end{itemize}

The inner pipeline computes:

\begin{align}
  \mathbf z_{\rm lab}(t) &= \mathsf{N}^{-1}\mathbf z_{\rm loops}(t),\\
  \mathbf z_\perp(t) &= \mathsf{P}_\perp \mathbf z_{\rm lab}(t),\\
  I(t) &= |\mathbf z_\perp|^2,\\
  A_p(t) &= \hat{\mathbf p}\cdot\mathbf z_\perp,\quad
  A_q(t) = \hat{\mathbf q}\cdot\mathbf z_\perp,
\end{align}
then forwards $(A_p, A_q)$ to \texttt{stokes.compute\_stokes()} which
returns the four Stokes parameters, the polarization fraction
$\sqrt{Q^2+U^2+V^2}/I$, the ellipticity $\frac12 \arcsin(V/I_p)$ in
degrees ($\pm 45^\circ$ at full circular polarization), and the
position angle $\frac12 \arctan2(U, Q)$.

A $2\times 2$ coherency on $(A_p, A_q)$ identifies the dominant-pol
eigenvector $u_{\rm dom}$, and
$z_{\rm dom}(t) = u_{\rm dom}^\top (A_p, A_q)$ provides an instantaneous
phase signal whose unwrapped, linearly-detrended residual gives both
the residual carrier frequency and a time-resolved frequency series via
\texttt{numpy.gradient}.  Everything is bundled into an
\texttt{AnalysisResult} dataclass.

\subsection{\texttt{triloop.multiband}}

Wraps band extraction and per-band analysis, with optional projection
onto the magnetoionic O/X modes.  Public functions:

\begin{itemize}\itemsep=2pt
\item \texttt{analyze\_all\_bands(cap, az\_deg, el\_deg,
      loops\_channels, bw\_hz, decim\_rate\_hz, rf\_bands, loops\_config,
      tx\_lat\_deg, tx\_lon\_deg, rx\_lat\_deg, rx\_lon\_deg,
      reflection\_height\_km)}
      -- calls \texttt{extract\_bands} once and runs
      \texttt{analyze\_z\_loops} on each band.  When all four TX/RX
      coordinates are supplied, also calls
      \texttt{exit\_point\_geometry} once and projects each band's
      perp-plane signal onto the predicted O/X modes (see
      Section~\ref{sec:magnetoionic-ref}).  Returns a dict mapping
      \texttt{rf\_hz} $\to$ \texttt{MultiBandResult} with fields
      \texttt{(rf\_hz, extraction, analysis, snrs\_db, mode)} where
      \texttt{mode} is a \texttt{ModeSeparation} or \texttt{None}.
\item \texttt{make\_multiband\_figure(results, out\_path)} -- plot a
      five-row, four-column matplotlib figure (one row per band) of
      intensity, instantaneous frequency, ellipticity, and polarization
      fraction.
\end{itemize}

%-----------------------------------------------------------------------
\section{Magnetoionic O/X mode separation}
\label{sec:magnetoionic-ref}
%-----------------------------------------------------------------------

\subsection{Physical motivation}

In the cold magnetised plasma of the ionosphere, an HF wave does not
propagate as a single mode --- it splits into two orthogonal
characteristic modes (the \emph{ordinary} O and \emph{extraordinary} X)
with slightly different refractive indices.  Each mode carries its own
amplitude and phase along the ray, and the polarization \emph{shape}
of each mode evolves smoothly with the local
$\theta = \angle(\hat k, \hat B)$.

At the ray's \emph{exit point} (where the wave is just leaving the
ionosphere on its way to the receiver), the local plasma frequency
$f_p \to 0$ so the mode polarization simplifies to a pure geometric
calculation in $(\theta, f_H, f)$.  From the exit point onward the
modes propagate adiabatically through vacuum to the antenna, with no
further polarization evolution.  This is why the exit point is the
right place to evaluate the mode shapes for receiver-side analysis.

\subsection{The Appleton-Hartree polarization parameter}

For a wave $\mathbf{E} = E_p\hat p_B + \rho E_p\hat q_B$ in the
right-handed perpendicular-plane basis $(\hat p_B, \hat q_B)$ with
$\hat p_B$ along the projection of $\hat B$, the cold-plasma result is
\begin{equation}
\rho_{O,X} \;=\; -iq \;\pm\; i\sqrt{q^2 + 1},
\qquad q \;=\; \frac{Y_T^2}{2\,Y_L\,(1-X)},
\end{equation}
with $Y \equiv f_H/f$, $Y_L = Y\cos\theta$, $Y_T = Y\sin\theta$,
$X = (f_p/f)^2$.  At the exit point $X = 0$.

Limiting cases (verified numerically by the sanity-check suite, see
Section~\ref{sec:magnetoionic-validation}):
\begin{itemize}\itemsep=2pt
\item $\theta \to 0$ (QL): $\rho_O \to +i$, $\rho_X \to -i$ -- pure
      RCP/LCP about $\hat k$.  Both modes are circular.
\item $\theta \to \pi/2$ (QT): $\rho_O \to 0$, $\rho_X \to -i\infty$
      -- two orthogonal linear polarizations along $\hat p_B$ and
      $\hat q_B$.
\item All $\theta$: the two modes are orthogonal under the Hermitian
      inner product, $\langle u_O, u_X\rangle = 0$, where
      $u_M = (1, \rho_M)/\sqrt{1+|\rho_M|^2}$.  Verified to numerical
      precision on $10^3$ random parameter draws.
\end{itemize}

\subsection{Sign and handedness conventions}

\begin{itemize}\itemsep=2pt
\item $\hat k$ is the Poynting direction --- source $\to$ receiver.
      Use \texttt{khat\_propagation(az\_src, el\_src)} or just negate
      what \texttt{az\_el\_to\_khat} returns.
\item $(\hat p_B, \hat q_B)$ is right-handed:
      $\hat p_B \times \hat q_B = \hat k$.
\item $\rho$ describes polarization relative to $\hat k$, not to
      $\hat B$.  In particular, flipping $\theta \to \pi - \theta$
      does \emph{not} negate $\rho$ in our formulation; we verified
      this both algebraically (Block 1 of the sanity-check suite) and
      numerically.  The handedness ``relative to $\hat B$'' is what
      transforms simply, but for matched-filter projection at the
      receiver the relevant frame is $\hat k$.
\item Texts differ on the O/X label at $\theta = 0$.  Our code matches
      the IEEE/PHaRLAP convention (O is the wave with $\rho = +i$,
      i.e.\ right-handed about $+\hat k$).  Since both modes appear in
      any real wave, the label is a bookkeeping choice; what matters
      is that the two outputs of \texttt{appleton\_hartree\_polarization}
      are orthogonal.
\end{itemize}

\subsection{Path geometry: entry, exit, and the symmetry that doesn't hold}

\texttt{exit\_point\_geometry(tx\_lat, tx\_lon, rx\_lat, rx\_lon, h\_km)}
returns the geometry at the ray's descending exit point: a
flat-Earth-approximated three-quarter-path location at altitude
$h_{\rm km}$, with $\hat k$ pointing down and toward the receiver in
local exit-ENU.  \texttt{entry\_point\_geometry} returns the same
quantities at the upgoing entry (quarter path, $\hat k$ pointing up
and toward the apex).  For paths beyond ~2200\,km a single F2 hop
will not fit; \texttt{multi\_hop\_geometry(..., n\_hops=N, ...)}
returns the last-hop \texttt{ExitGeometry} of an $N$-hop ground-bounce
path, which is the wave that physically arrives at the receiver.

\paragraph{Apparent angle of arrival.}  The
\texttt{ExitGeometry} dataclass carries three AoA-relevant fields:
\texttt{el\_rx\_deg} (the flat-Earth elevation, used internally for
the $\hat k$ direction), \texttt{el\_rx\_spherical\_deg} (the more
accurate spherical-Earth value via Davies eq.\ 4.5), and
\texttt{az\_rx\_to\_src\_deg} (the great-circle initial bearing
from RX to TX).  The flat- vs spherical-Earth elevation difference
is $\sim h/(2R_e) \cdot $\,(range-dependent factor) — around
2--4$^\circ$ for 800--1400\,km hops at $h = 250$\,km — and the user
should always quote the spherical value for antenna pointing or
operator-facing reports.  See Section~\ref{sec:worked-paths-ref} for
the FC $\to$ APO and FC $\to$ Boston worked tables.

\paragraph{The symmetry test we expected and the one that actually
holds.}  Naively, $\theta_{\rm entry} + \theta_{\rm exit}$ should equal
$180^\circ$, since $\hat k$ flips its vertical component while $\hat B$
stays roughly the same.  The sanity-check suite found this fails: the
sum is closer to 234$^\circ$ for the FC$\to$APO path.  Working it out
algebraically: with $\hat k_{\rm entry} = (a, b, +c)$ and
$\hat k_{\rm exit} = (a, b, -c)$, $\cos\theta_{\rm entry}$ and
$\cos\theta_{\rm exit}$ differ only in the $\hat B$-vertical term.
Their arccosines sum to $180^\circ$ \emph{only} when the horizontal
components of $\hat k$ and $\hat B$ are perpendicular ---
i.e.\ when the path's horizontal direction is east-west and $\hat B$
has no horizontal east component.  For paths that aren't due-east
(FC$\to$APO is mostly N-S, with $\hat B$ also pointing mostly N), the
expected ``180$^\circ$'' identity does not hold.  The k-vector
identity $\hat k_{\rm entry,horiz} = \hat k_{\rm exit,horiz}$ and
$\hat k_{\rm entry,vert} = -\hat k_{\rm exit,vert}$ does hold to
machine precision.

\subsection{Path-integrated Faraday rotation}

\texttt{integrated\_faraday\_rotation\_rad(B\_cos\_theta\_avg\_T,
TEC\_electrons\_per\_m2, f\_hz)} returns
\begin{equation}
\Omega \;=\; \frac{e^3}{8\pi^2\varepsilon_0 m_e^2 c}\;
        \frac{\langle B\cos\theta\rangle\,\mathrm{TEC}}{f^2}
        \;\approx\; 2.36 \times 10^4
        \;\frac{\langle B\cos\theta\rangle[\mathrm{T}]\;\mathrm{TEC}[e/m^2]}{f^2[\mathrm{Hz}]}.
\end{equation}
At 30\,TECU on the FC$\to$APO path with
$\langle B\cos\theta\rangle \approx 25\,\mu$T, this gives $\sim 4500$
turns at 2.5\,MHz and $\sim 70$ turns at 20\,MHz.  The $f^{-2}$ scaling
is verified to numerical precision in the sanity-check suite.

\subsection{Receiver-side projection}

The connection from antenna data to mode amplitudes:
\begin{enumerate}\itemsep=2pt
\item \texttt{analyze\_z\_loops} produces $A_p(t), A_q(t)$ in the
      receiver's horizontal-first $(\hat p_R, \hat q_R)$ basis.
\item \texttt{multiband.\_project\_onto\_modes} computes the angle
      $\alpha$ between $\hat p_R$ and the magnetoionic $\hat p_B$ in
      the perp-plane, then rotates $(A_p, A_q) \to (A_{p_B}, A_{q_B})$.
\item Matched-filter:
      $a_O(t) = u_O^\dagger\,(A_{p_B}, A_{q_B})$,
      $a_X(t) = u_X^\dagger\,(A_{p_B}, A_{q_B})$.
\end{enumerate}
The differential phase $\Delta\varphi(t) = \arg a_O - \arg a_X$,
unwrapped, is the time-resolved analogue of the path-integrated
Faraday rotation.

\subsection{Sanity checks performed}
\label{sec:magnetoionic-validation}

The full validation suite runs eight blocks of checks, of which 21/23
named assertions pass.  The two that ``fail'' are:
\begin{itemize}\itemsep=2pt
\item ``$\rho_O$ at $\theta = \pi$ should equal $-\rho_O$ at $\theta = 0$.''
      It doesn't, because $\rho$ tracks polarization relative to
      $\hat k$, and flipping $\theta$ doesn't change the sign of $q$.
      This was a faulty expectation, not a code bug.
\item ``$\theta_{\rm entry} + \theta_{\rm exit} \approx 180^\circ$.''
      As discussed above, this only holds in special path geometries.
      The k-vector identity that \emph{does} always hold passes.
\end{itemize}
The passing checks include: AH limits at $\theta=0, \pi/2$;
mode orthogonality on $10^3$ random parameter draws; energy
conservation $|a_O|^2 + |a_X|^2 = |E|^2$ to $10^{-15}$; IGRF reality
check (mid-latitude US gives $|B| \approx 50\,\mu$T,
inclination $\approx 62^\circ$, declination $\approx 9^\circ$);
flat-Earth elevation within $2.5^\circ$ of the spherical-Earth value;
range haversine match to 0.2\%; numerical robustness at
$\theta = 0, \pi, \pi/2 \pm \epsilon$, $f = f_H$, $f \ll f_H$,
$f \gg f_H$.

\paragraph{When to use PHaRLAP instead.}  This module's flat-mirror
geometry is fine for first-pass mode-polarization signature work --
``are we in QL or QT?'' is robust against the geometric error.  For
quantitative work the right tool is PHaRLAP (Cervera, DSTG/U.\
Adelaide), which traces the full 3-D Haselgrove ray equations through
an IRI ionosphere with IGRF.  See \texttt{triloop.pharlap\_predict}
(planned) for the offline-prediction interface.

\subsection{Worked paths: WWV $\to$ APO and WWV $\to$ Boston}
\label{sec:worked-paths-ref}

These two paths cover the typical 1-hop and multi-hop regimes for
mid-latitude HF polarimetry.  Both receivers listen to WWV's
transmitter at Fort Collins, CO ($40.68^\circ$\,N, $-105.04^\circ$\,W).

\paragraph{FC $\to$ APO  (1-hop, 881\,km).}
$\theta_{\rm exit} = 70^\circ$, ray elevation at RX is $27.1^\circ$
(spherical) / $29.6^\circ$ (flat-Earth), great-circle bearing
$4.3^\circ$ (effectively due N).  Mode axial ratio runs from 0.52 at
2.5\,MHz to 0.93 at 25\,MHz --- modes are moderately linear at low
$f$ and near-circular at high $f$.  Cross-pol discrimination grows
from +10\,dB at 2.5\,MHz to +29\,dB at 25\,MHz.

\paragraph{FC $\to$ Boston (Harvard, 2812\,km, 2-hop).}
Last-hop range $1406$\,km, $\theta_{\rm exit} = 55^\circ$, ray
elevation at RX is $16.0^\circ$ (spherical) / $19.6^\circ$
(flat-Earth), great-circle bearing $277.7^\circ$ (effectively due W).
Mode axial ratio runs from 0.72 at 2.5\,MHz to 0.97 at 25\,MHz ---
the smaller exit-point $\theta$ pushes the modes more circular than
APO at every band.  XPD reaches $+25$\,dB by 5\,MHz vs.\ 10\,MHz at APO.

\paragraph{Geometric AR is not the RCP/LCP energy split.}  The
$P_{\rm dom}/P_{\rm tot}$ column in the user-manual tables uses the
identity $P_{\rm dom}/P_{\rm tot} = (1+r)^2/(2(1+r^2))$ for axial
ratio $r$.  The relationship is quadratic:  axial ratio 0.5 (visually
``moderately linear'') gives a 90\,\%/10\,\% split, not 50\,\%/50\,\%.
The helper \texttt{circular\_energy\_fractions(axial\_ratio)} returns
the pair, and \texttt{cross\_pol\_discrimination\_db(axial\_ratio)}
returns the dB ratio, useful for receiver design or for predicting
how much power a circular-only antenna will couple from a given
band.

The user manual's \S Worked paths gives the full 5-band tables.
Reproduce them via:
\begin{lstlisting}[style=py]
from triloop import multi_hop_geometry, modes_at_exit
g_apo = multi_hop_geometry(40.6796, -105.0411, 32.7803, -105.8200,
                           n_hops=1, h_km=250.0)
g_bos = multi_hop_geometry(40.6796, -105.0411, 42.3744, -71.1169,
                           n_hops=2, h_km=250.0)
for f_MHz in [2.5, 5, 10, 15, 20]:
    m_apo = modes_at_exit(g_apo, f_MHz * 1e6)
    m_bos = modes_at_exit(g_bos, f_MHz * 1e6)
    print(f"{f_MHz:5.1f} MHz: APO AR={m_apo.axial_ratio_O:.3f}, "
          f"Boston AR={m_bos.axial_ratio_O:.3f}")
\end{lstlisting}

%-----------------------------------------------------------------------
\section{Direction finding}
%-----------------------------------------------------------------------

\subsection{\texttt{triloop.direction}}

Public functions, all operating on a stacked lab-frame complex array
$\mathsf{Z}_{\rm lab}\in\mathbb{C}^{3\times N}$:

\begin{itemize}\itemsep=2pt
\item \texttt{coherency\_matrix(Z\_lab)} returns
      $\mathsf{C} = \tfrac{1}{N}\mathsf{Z}_{\rm lab}\mathsf{Z}_{\rm lab}^H$,
      the $3\times 3$ Hermitian coherency.
\item \texttt{estimate\_direction\_from\_z\_lab(Z\_lab)} returns the
      smallest-eigenvalue eigenvector of $\mathsf{C}$ as
      $\hat{\mathbf k}_{\rm est}$, plus the polarization basis
      ($\hat{\mathbf p}$, $\hat{\mathbf q}$ from the two largest
      eigenvectors), the eigenvalues (descending), and a scalar
      \texttt{null\_strength} $\lambda_3/\lambda_1$.  This is the
      \emph{coarse} direction estimate; small \texttt{null\_strength}
      means the rank-1 trap (\S~\ref{sec:rank1ref}) is broken and
      the estimate is trustworthy.
\item \texttt{perp\_residual\_energy(Z\_lab, khat)} computes
      $\langle |\hat{\mathbf k}\cdot \mathbf z_{\rm lab}|^2 \rangle$,
      the quantity minimised by direction finding.
\item \texttt{null\_search(Z\_lab, az0, el0, half\_width\_deg, n\_points)}
      performs a 2-D grid sweep around $(az_0, el_0)$ and returns the
      minimiser plus the full residual map.
\item \texttt{parabolic\_refine(R, az\_grid, el\_grid)} fits a
      parabola to the $3\!\times\!3$ neighbourhood of the grid
      minimum to give a sub-grid direction.
\item \texttt{lock\_and\_analyze(...)} -- end-to-end entry point used
      by the \texttt{locate} CLI.
\end{itemize}

\subsection{Why find the null instead of the maximum?}
\label{sec:rank1ref}

The direction-formed power
$P(\hat{\mathbf k}) = \langle|\mathbf B_\perp(\hat{\mathbf k})|^2\rangle$
varies as $\frac12 + \frac12\cos^2\beta$ near the true direction
$\hat{\mathbf k}_{\rm true}$ (a shallow plateau).  The
\emph{perpendicular} residual
$Q(\hat{\mathbf k}) = \langle|\hat{\mathbf k}\cdot\mathbf B|^2\rangle$
varies as $\sin^2\beta\sim\beta^2$ near the same direction (a sharp
parabolic minimum).  Localising a parabolic minimum is far more
precise than a cosine-squared maximum, so we always lock on the null
rather than the peak.

For purely linearly polarized signals from a single direction,
$\mathsf{C}$ is rank 1 and the null is a 2-D plane rather than a
point.  In that regime \texttt{estimate\_direction\_from\_z\_lab}
returns \emph{a} direction (the smallest-eigenvalue one within
floating-point noise), but it is not physically meaningful; trustworthy
localisation requires Faraday rotation, a circularly polarized source,
or triangulation.

\subsection{\texttt{triloop.beamform}}

Older, simpler beamforming sweep used by some analysis notebooks.
Two functions:

\begin{itemize}\itemsep=2pt
\item \texttt{beamform\_grid(z\_loops, loops\_config, az\_grid\_deg,
      el\_grid\_deg)} -- compute time-averaged
      $P(\varphi, \theta) = \langle|\mathbf B_\perp(\hat{\mathbf k}_{\varphi,\theta})|^2\rangle$
      over a 2-D grid.
\item \texttt{best\_direction(P, az\_grid, el\_grid)} -- return the
      $\arg\max$ as $(\varphi, \theta)$.
\end{itemize}

\subsection{\texttt{triloop.stokes}}

A single function: \texttt{compute\_stokes(A\_p, A\_q)} returning a
dict with \texttt{I, Q, U, V, pol\_fraction, ellipticity\_deg,
position\_angle\_deg}.  Definitions:
$I = |A_p|^2 + |A_q|^2$,
$Q = |A_p|^2 - |A_q|^2$,
$U = 2\Re\{A_p^*A_q\}$,
$V = 2\Im\{A_p^*A_q\}$.

%-----------------------------------------------------------------------
\section{Reporting and visualization}
%-----------------------------------------------------------------------

\subsection{\texttt{triloop.view}}

Static QC PNG generator used by \texttt{triloop view}.
\texttt{make\_view\_figure(cap, rf\_bands, bw\_hz, out\_path)}
produces a single matplotlib figure with three rows: a full-Nyquist
PSD with each RF band annotated at its alias position, per-band PSD
zooms (with the displayed Δf-axis flipped for inverted bands so the
operator reads frequencies the right way), a decimated time-domain
trace of every channel, and the auto-range probe table.

\subsection{\texttt{triloop.browse}}

Interactive single-file HTML browser.  Top-level entry points:

\begin{itemize}\itemsep=2pt
\item \texttt{build\_browser\_html(cap, az\_deg, el\_deg,
      loops\_channels, bw\_hz, decim\_rate\_hz, n\_anim\_frames,
      title)} -- returns a self-contained HTML string.
\item \texttt{write\_browser(cap, out\_path, **kw)} -- save it to disk.
\end{itemize}

The page uses Plotly (loaded once via CDN; if you need a fully offline
build, change the \texttt{include\_plotlyjs="cdn"} argument to
\texttt{True} and the page will inline the entire library at the cost
of $\sim 3$\,MB extra).  Layout: tab bar at the top with
\textit{Overview}, \textit{Time series}, and one tab per RF band.
The per-band tab has four panels: PSD zoom, intensity time history
(linear + togglable dB), intensity CDF with P10/P1 fade-depth markers,
and an animated polarization ellipse with Play/Pause and a frame
slider.

The animation is implemented as Plotly frames; each frame draws a
sliding window of $(\Re A_p, \Re A_q)$ samples centred on a different
slow-time epoch.  The window length is chosen automatically to be a
few cycles of the residual carrier-phase variation at the decimated
rate.

%-----------------------------------------------------------------------
\section{Module roll call}
%-----------------------------------------------------------------------

This section lists every Python file in both packages and what each
one provides.  Treat it as a table of contents for the source.

\subsection{\textsf{picoacq}}

\begin{tabular}{p{0.21\linewidth}p{0.74\linewidth}}
\toprule
\textbf{Module} & \textbf{Purpose} \\\midrule
\texttt{\_\_init\_\_.py} & package marker; no public API of its own. \\
\texttt{cli.py}   & Click-based CLI; subcommands \texttt{capture} and
                    \texttt{monitor}.  Reads
                    \texttt{--rate / --duration / --channels /
                    --range / --auto-range / --headroom / --probe /
                    --rf-bands / --simulate / --sim-*} options and
                    forwards them to \texttt{recorder.capture()}. \\
\texttt{recorder.py} & Top-level \texttt{capture()} orchestrator with
                    \texttt{\_check\_buffer()} guard,
                    \texttt{auto\_range()} probe, range chooser
                    \texttt{\_pick\_range()}, and the
                    \texttt{\_try\_real\_capture()} hardware path
                    (with simulator fallback). \\
\texttt{ps5444a.py} & PicoSDK 5000a wrapper: \texttt{open\_5444()},
                    \texttt{configure\_channel()},
                    \texttt{get\_timebase()}, \texttt{acquire\_block()}.
                    Plus \texttt{\_ensure\_pico\_dyld\_path()} for
                    macOS framework search. \\
\texttt{simulator.py} & \texttt{simulate\_capture()} -- 3-loop
                    no-hardware fallback compatible with the
                    \texttt{recorder} return shape. \\
\texttt{alias.py}    & RF~$\leftrightarrow$~baseband mapping under
                    bandpass undersampling.  Public:
                    \texttt{Alias} dataclass,
                    \texttt{alias\_of()},
                    \texttt{aliases\_for\_bands()},
                    \texttt{rf\_for\_baseband()}.  See
                    Section~\ref{sec:alias}. \\
\bottomrule
\end{tabular}

\subsection{\textsf{triloop}}

\begin{tabular}{p{0.21\linewidth}p{0.74\linewidth}}
\toprule
\textbf{Module} & \textbf{Purpose} \\\midrule
\texttt{\_\_init\_\_.py} & re-exports the public API (see
                    \textit{User Manual} for the full list). \\
\texttt{cli.py}   & Click-based CLI; subcommands
                    \texttt{view, analyze-multi, browse, analyze,
                    locate, batch, summary, simulate, notebook}. \\
\texttt{io\_hdf5.py}  & \texttt{read\_capture()} and
                    \texttt{write\_capture()} for the HDF5 v0.1 layout
                    described in \S\ref{sec:fileformat-ref}. \\
\texttt{config.py}    & \texttt{default\_loops\_config()} (cube-vertex
                    geometry) and \texttt{validate\_loops\_config()}
                    schema check. \\
\texttt{geometry.py}  & \texttt{az\_el\_to\_khat()},
                    \texttt{build\_N\_matrix()},
                    \texttt{perp\_projector()},
                    \texttt{perp\_orthonormal\_basis()},
                    \texttt{LOOP\_NORMALS\_DEFAULT}. \\
\texttt{extract.py}   & Single-band complex-baseband extractor:
                    \texttt{extract\_complex\_baseband()},
                    \texttt{extract\_three\_loops()}. \\
\texttt{bands.py}     & Multi-band extractor:
                    \texttt{extract\_bands()},
                    \texttt{BandExtraction} dataclass. \\
\texttt{stokes.py}    & \texttt{compute\_stokes(A\_p, A\_q)} ->
                    Stokes parameters and derived quantities. \\
\texttt{analyze.py}   & Top-level \texttt{analyze()} and inner
                    \texttt{analyze\_z\_loops()}; \texttt{AnalysisResult}
                    dataclass. \\
\texttt{multiband.py} & \texttt{analyze\_all\_bands()} (extract +
                    analyze for every band, with optional O/X mode
                    separation when TX/RX coordinates are given) plus
                    \texttt{ModeSeparation} and \texttt{MultiBandResult}
                    dataclasses; private \texttt{\_project\_onto\_modes()}
                    matched-filter helper; \texttt{make\_multiband\_figure()}. \\
\texttt{magnetoionic.py} & Cold-plasma O/X mode polarization:
                    \texttt{appleton\_hartree\_polarization()},
                    \texttt{mode\_unit\_vectors()},
                    \texttt{mode\_ellipticity\_deg()},
                    \texttt{mode\_orientation\_deg()},
                    \texttt{mode\_axial\_ratio()},
                    \texttt{circular\_energy\_fractions()} (RCP/LCP
                    energy split from axial ratio),
                    \texttt{cross\_pol\_discrimination\_db()},
                    \texttt{khat\_propagation()},
                    \texttt{igrf\_b\_hat\_enu()} (with pyIGRF or
                    fallback approximation),
                    \texttt{haversine\_km()},
                    \texttt{initial\_bearing\_deg()},
                    \texttt{spherical\_elevation\_at\_rx\_deg()},
                    \texttt{gc\_intermediate\_point()},
                    \texttt{exit\_point\_geometry()} (returns flat-
                    Earth and spherical-Earth elevation, plus
                    great-circle bearing),
                    \texttt{entry\_point\_geometry()},
                    \texttt{multi\_hop\_geometry()} (last-hop
                    geometry for n-hop F2 ground-bounce paths),
                    \texttt{modes\_at\_exit()},
                    \texttt{integrated\_faraday\_rotation\_rad()};
                    \texttt{ExitGeometry, EntryGeometry, ModeAtExit}
                    dataclasses. \\
\texttt{direction.py} & Coherency-based direction finding:
                    \texttt{coherency\_matrix()},
                    \texttt{estimate\_direction\_from\_z\_lab()},
                    \texttt{perp\_residual\_energy()},
                    \texttt{null\_search()},
                    \texttt{parabolic\_refine()},
                    \texttt{lock\_and\_analyze()}. \\
\texttt{beamform.py}  & Older/simpler grid-sweep:
                    \texttt{beamform\_grid()},
                    \texttt{best\_direction()}. \\
\texttt{view.py}      & Static QC PNG: \texttt{make\_view\_figure()}. \\
\texttt{browse.py}    & Interactive Plotly HTML browser:
                    \texttt{build\_browser\_html()},
                    \texttt{write\_browser()}, plus the helpers
                    \texttt{\_channel\_psd()},
                    \texttt{\_band\_zoom\_spectrum()},
                    \texttt{\_smooth\_decimate()},
                    \texttt{\_cdf()},
                    \texttt{\_build\_pol\_animation()}. \\
\bottomrule
\end{tabular}

%-----------------------------------------------------------------------
\section{Tracing a capture through the code}
%-----------------------------------------------------------------------

To consolidate the picture above, here are the call paths for the
three new commands:

\paragraph{\texttt{triloop view cap.h5}.}
\begin{enumerate}\itemsep=2pt
\item \texttt{cli.view\_cmd} parses options.
\item \texttt{io\_hdf5.read\_capture(cap.h5)} -> dict.
\item \texttt{view.make\_view\_figure(cap, rf\_bands, bw\_hz, out\_path)}.
\item Inside \texttt{make\_view\_figure}: per-channel PSD via
      \texttt{\_channel\_psd}; per-band zoom via
      \texttt{\_band\_zoom\_spectrum} (calls
      \texttt{picoacq.alias.alias\_of} for each band centre);
      time-domain decimated; probe table from
      \texttt{capture\_settings.auto\_range\_probe}.
\end{enumerate}

\paragraph{\texttt{triloop analyze-multi cap.h5 --az 9 --el 35}.}
\begin{enumerate}\itemsep=2pt
\item \texttt{cli.analyze\_multi\_cmd} parses options.
\item \texttt{io\_hdf5.read\_capture(cap.h5)} -> dict.
\item \texttt{multiband.analyze\_all\_bands(cap, az, el, ...)}:
\begin{enumerate}\itemsep=1pt
\item \texttt{bands.extract\_bands(cap, ...)} which for each band uses
      \texttt{picoacq.alias.alias\_of} to find $f_{\rm bb}$, then mixes
      and decimates each channel via \texttt{\_extract\_one\_channel}.
\item For each band: \texttt{\_per\_loop\_snr\_from\_baseband}, then
      \texttt{analyze.analyze\_z\_loops(t, Z\_loops, f\_rf, ...)}.
\end{enumerate}
\item \texttt{multiband.make\_multiband\_figure(results, out\_png)}.
\item Per-band JSON summary written to \texttt{out\_json}.
\end{enumerate}

\paragraph{\texttt{triloop browse cap.h5}.}
\begin{enumerate}\itemsep=2pt
\item \texttt{cli.browse\_cmd} parses options.
\item \texttt{io\_hdf5.read\_capture(cap.h5)} -> dict.
\item \texttt{browse.write\_browser(cap, out\_html, ...)} which calls
      \texttt{build\_browser\_html(cap, ...)}:
\begin{enumerate}\itemsep=1pt
\item Calls \texttt{multiband.analyze\_all\_bands} (same path as
      above) to populate per-band intensity, ellipticity, and
      $A_p / A_q$ time series.
\item Builds per-tab Plotly figures: overview PSD, time series, and
      one 4-panel tab per band.
\item For each band assembles a polarization-ellipse animation via
      \texttt{\_build\_pol\_animation()} (Plotly frames + slider).
\item Renders the page using \texttt{plotly.io.to\_html()} for each
      figure into a tabbed HTML scaffold.
\end{enumerate}
\end{enumerate}

%-----------------------------------------------------------------------
\section{Testing and verification}
%-----------------------------------------------------------------------

The test suite under \texttt{triloop/tests/} exercises the analysis
pipeline against the simulator at known polarizations and directions.
Useful commands during development:

\begin{lstlisting}[style=shell]
# Create a synthetic 4-channel capture with all 5 WWV bands aliased
PYTHONPATH=triloop:picoacq python3 -m picoacq.cli capture \
    --simulate --duration 0.5 -o sim.h5

# Run the multi-band analysis and confirm pol_frac and ellipticity
PYTHONPATH=triloop:picoacq python3 -m triloop.cli analyze-multi sim.h5 \
    --az 9 --el 35

# Render the interactive browser
PYTHONPATH=triloop:picoacq python3 -m triloop.cli browse sim.h5
\end{lstlisting}

For the end-to-end ``does the alias arithmetic come out right'' check
on a synthetic 4-channel 15.625\,MS/s capture, see
\texttt{triloop/triloop/bands.py} (specifically \texttt{extract\_bands}
will raise \texttt{ValueError} if the requested per-band bandwidth
overlaps the Nyquist edge or another band's alias, which is the
nearest thing to an automatic correctness check).

\end{document}
